Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies

### 2019-06-12 man/ecosystems/ms_ecosystems.Rmd
update

## Manuscript highlights
* H1
* H2
* H3

@@

@@

$$\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$$
where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except $i$ year); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period. Each pixel was categorized according the EVI standardized anomalies as "greening" (EVI standardized anomalies greater than + 1), "browning" (EVI standardized anomalies less than - 1) or "no changes" (EVI standardized anomalies between − 1 and  + 1) [@Samanta2012].

Rather than other vegetation indices like the NDVI, $EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual ($r$ = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88). For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale. The nonparametric Mann–Kendall test was applied.

@@

In addition, the forest history of San Juan and Cáñar sites was inferred from a detailed analysis of historical land-use changes. For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest: land-uses, forest managment actions, mining, wildfires, etc. Several documentary sources were reviewed: historical documents and maps [*e.g.* @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2014]; traditional irrigation ditches [@RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@JimenezOlivencia2015; @MorenoLlorca2016].

@@

To evaluate the effects of drougth events on ecosystem resilience (using greeennes data) and tree individual resilience (using BAI growth data) we used resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*). The Resistance index, estimated as the ratio between the performance during and before the disturbance ($Rt=Drought / PreDrought$), quantifies the severity of the impact of the disturbance in the year it occurred. The Recovery index, computes as the ratio between performance after and during disturbance ($Rc = PostDrought / Drought$), represents the ability to recover from disturbance relative to its severity. Finally, the Resilience index ($Rs = PostDrought / PreDrought$) is the capacity to reach pre-disturbance performance levels. 
The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event. 2005 and 2012 were considered as single drought events. The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively. A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S6) and this length was used in other studies [*e.g.* @Gazol2017]. To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1950 and compared them with drought severity (Figure 6, Table S3).

@@

For each of the three resilience indices studied, robust two-way ANOVAs were used to test for differences between drought events (2005 and 2012) and the two studied oak populations (northern and southern aspects). These tests were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012]. When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers. Post-hoc differences were assessed pairwise using a similar boostrap test. All the robust ANOVA and post-hoc tests were carried out using the `WRS2` package. The level of significance was set to 0.05 and adjusted for multiple comparisons.

@@

A remarkably result is the higher values of resilience for tree-growth obtained during the worst and more severe drought events in our study area (1995 and 1999) (Table S3; Figure 6). A plausible explanation could be the water availability previously and after the drought event. For instance, cumulative precipitation of the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern. In fact, the first months of 1996 recorded the highest cumulative rainfall of the past century for our southern site (Cáñar) [@MesaTorres2009]. This could explain the higher values of recovery found for 1995 (Figure 6).

@@

When we explored the chracteristics of the southern sites, we firstly observed a difference in elevation. @Bhuyan2017, in a recent work across the European continent, found a clearly effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive. The higher values of tree growth that we observed in CA-High site could be related with lower water stress at high-elevation sites, since moisture availability is a key factor limiting tree growth for *Q. pyrenaica* [@GeaIzquierdo2014]. Soil analyses along an elevational gradient in this location revealed higher content of available water for forest soils located on high elevation [@CoboDiaz2017]. 
In this sense, a remarkable observation is the presence of traditional irrigation ditches which could provide an additional water supply. In most of the high-mountain watersheds of Sierra Nevada, there are systems of historical irrigation channels, know as *acequias de careo*, that was used since Middle Age to cultivated these valleys [@MartinCivantos2014]. These ditches run through the hillsides of valleys releasing water through several points that filtrate the water to recharge the aquifer and then irrigates the foot of the slopes [@MartinCivantos2014; @Jodar2017]. Several works have recently described the network of traditional ditches for our southern sites (Cáñar), with detailed information about water use [@RuizRuiz2017] and its hydrological functioning [@MartinMontanes2015]. There is a ditch (*Acequia de la Era Alta*) located uphill the CA-High site, which functions from March to June [@RuizRuiz2017]. It brings water from snowmelt and from a nearby stream, soaking the *Q. pyrenaica* forests. This traditional system could supply an extra of water that could be used by the trees located downstream of these channels. This extra of water is particularly important for trees of this dry-rear edge and could explain the higher values of BAI for trees located in CA-High site.

>>>>

## Manuscript highlights
* H1
* H2 %>% %>% 
* H3

@@

Mira la biblio de regino y esto https://www.researchgate.net/publication/327321775_Drought_Sensitiveness_on_Forest_Growth_in_Peninsular_Spain_and_the_Balearic_Islands

@@

$$\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$$
where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except $i$ year); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period. Each pixel was categorized according the EVI standardized anomalies as "greening" ($EVI~sa > 1$), "browning" ($EVI~sa <- 1$) or "no changes" ($-1 > EVI~sa > 1$)[@Samanta2012].

Rather than other vegetation indices like the NDVI, $EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual ($r$ = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88). For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale, using Mann–Kendall nonparametric test.

@@

In addition, the forest and managment history of our sampling sites was inferred from a detailed analysis of historical land-use changes. For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest: land-uses, forest managment actions, mining, wildfires, etc. Several documentary sources were reviewed: historical documents and maps [*e.g.* @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2016]; traditional irrigation ditches [@RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@JimenezOlivencia2015; @MorenoLlorca2016].

@@

To evaluate the effects of drougth events on ecosystem resilience (using greeennes data) and tree individual resilience (using BAI growth data) we used resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*). The Resistance index, estimated as the ratio between the performance during and before the disturbance ($Rt=Drought / PreDrought$), quantifies the severity of the impact of the disturbance in the year it occurred. The Recovery index, computes as the ratio between performance after and during disturbance ($Rc = PostDrought / Drought$), represents the ability to recover from disturbance relative to its severity. Finally, the Resilience index ($Rs = PostDrought / PreDrought$) is the capacity to reach pre-disturbance performance levels. 
The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event. 2005 and 2012 were considered as single drought events. The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively. A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S6) and this length was used in other studies [*e.g.* @Gazol2018]. To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1950 and compared them with drought severity (Figure 6, Table S3).

@@

For each of the three resilience indices studied, robust two-way ANOVAs were used to test for differences between drought events (2005 and 2012) and the studied oak populations (northern and southern aspects). These tests were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012]. When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers. Post-hoc differences were assessed pairwise using a similar boostrap test. All the robust ANOVA and post-hoc tests were carried out using the `WRS2` package. The level of significance was set to 0.05 and adjusted for multiple comparisons.

@@

A remarkably result is the higher values of resilience for tree-growth obtained during the worst and more severe drought events in our study area (1995 and 1999) (Table S3; Figure 6). A plausible explanation could be the water availability previously and after the drought event. For instance, cumulative precipitation of the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern. In fact, the first months of 1996 recorded the highest cumulative rainfall of the past century for our southern site (Cáñar). This could explain the higher values of recovery found for 1995 (Figure 6).

@@

When we explored the chracteristics of the southern sites, we firstly observed a difference in elevation. @Bhuyan2017, in a recent work across the European continent, found a clearly effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive. The higher values of tree growth that we observed in CA-High site could be related with lower water stress at high-elevation sites, since moisture availability is a key factor limiting tree growth for *Q. pyrenaica* [@GeaIzquierdo2014]. Soil analyses along an elevational gradient in this location revealed higher content of available water for forest soils located on high elevation [@CoboDiaz2017].

```{r eval=FALSE, echo=FALSE}
# In this sense, a remarkable observation is the presence of traditional irrigation ditches which could provide an additional water supply. In most of the high-mountain watersheds of Sierra Nevada, there are systems of historical irrigation channels, know as *acequias de careo*, that was used since Middle Age to cultivated these valleys [@MartinCivantos2014]. These ditches run through the hillsides of valleys releasing water through several points that filtrate the water to recharge the aquifer and then irrigates the foot of the slopes [@MartinCivantos2014; @Jodar2017]. Several works have recently described the network of traditional ditches for our southern sites (Cáñar), with detailed information about water use [@RuizRuiz2017] and its hydrological functioning [@MartinMontanes2015]. There is a ditch (*Acequia de la Era Alta*) located uphill the CA-High site, which functions from March to June [@RuizRuiz2017]. It brings water from snowmelt and from a nearby stream, soaking the *Q. pyrenaica* forests. This traditional system could supply an extra of water that could be used by the trees located downstream of these channels. This extra of water is particularly important for trees of this dry-rear edge and could explain the higher values of BAI for trees located in CA-High site. 
```

### 2019-06-12 man/figures/resilience_long_term.Rmd
update

sdf <- df %>% 
  filter(samps >= 10) %>% 
  group_by(disturb_year) %>% 
  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
  inner_join(dy, by=c("disturb_year" = "maxyear"))

@@

>>>>

sdf <- df %>% 
  filter(samps >= 10) %>% 
  group_by(disturb_year) %>% 
  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
  inner_join(dy, by=c("disturb_year" = "maxyear")) 
```

@@

```{r}
library("purrr")

sdfsite <- df %>% 
  filter(samps >= 10) %>% 
  group_by(disturb_year, site) %>% 
  filter(disturb_year > 1940) %>% 
  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
  inner_join(dy, by=c("disturb_year" = "maxyear"))

plotrelsite <- function(df, v, x, label, site, coordX_eq, coordY_eq, titulo, ...){
  require(ggplot2)
  require(ggrepel)
  require(purrr)

  eqns <- df %>% split(.$site) %>% 
    map(~ lm(formulita, data = .)) %>%
    map(lm_eqn) %>% 
    do.call(rbind, .) %>%
    as.data.frame() %>%
    set_names("equation") %>%
    mutate(site = rownames(.))

  ggplot(df, aes_string(y=paste0(v, "_mean_mean"), x=x, label=label, 
                        colour = site, fill = site)) + 
    geom_smooth(method='lm', se=FALSE) + 
    geom_point(size=1.5, pch = 21) + 
    theme_bw() + 
    geom_text_repel(nudge_x = 0.5, size = 3) + 
    annotate("text", x = coordX_eq, y = coordY_eq, 
             label=lm_eqn(lm(formulita, df)), parse=TRUE, 
             size = 3, fontface = "bold", hjust="left") +
    ggtitle(titulo) + 
    theme(plot.title = element_text(size = 11, face = "bold.italic"),
          panel.grid = element_blank(), 
          axis.text.y = element_text(size=9),
          axis.text.x = element_text(size=9),
          axis.title = element_text(size = 10)) +
    geom_text_repel(data = eqns, aes(x = -Inf, y = Inf,label = equation), parse = TRUE, segment.size = 0)

plotrelsite(sdfsite, v = "rs", x = 'd_severity', label = 'disturb_year',
                 site = "site",
                 coordX_eq = 6, coordY_eq = 0.5, titulo = "Rt") +
  xlab('') + ylab('') + 
  theme(plot.margin = margin(r=-0.7, l=0, t=0, b=0, "pt"))
